# Chapter 6: Analysis of longitudinal data

Data creation R script:   
https://github.com/tnissinen/IODS-project/blob/master/data/meet_and_repeat.R

### Loading and checking the structure of the data

```{r readdata_rats,echo=TRUE,results='hide',message=FALSE,warning=FALSE}

# access libraries
library(dplyr)
library(tidyr)
library(ggplot2)

```

We use the RATS dataset. It contains data from a nutrition study conducted in three groups of rats. The three groups were put on different diets, and each animal’s body weight (grams) was recorded repeatedly.

```{r}
# load the RATS data
RATS <- read.csv("~/IODS-project/data/rats_long.csv")

# explore the dataset
str(RATS)
summary(RATS)
dim(RATS)

# Factor variables ID and Group
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

```

176 observations with 6 columns in the dataset.


### Graphical Displays of Longitudinal Data

Plot the weight values for all the 16 rats

```{r fig_rats, fig.path="figures/", fig.dim=c(7,5), results='hide', message=FALSE, warning=FALSE}

# Draw the plot
ggplot(RATS, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATS$Weight), max(RATS$Weight)))

```

Individual weight profiles by group. The weight seems to be increasing over time for all the rats.



```{r fig_rats2, fig.path="figures/", fig.dim=c(7,5), results='hide', message=FALSE, warning=FALSE}

# Standardise the variable RATS
RATS <- RATS %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

# Plot again with the standardised RATS
ggplot(RATS, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")


```

Individual weight profiles for RATS data after standardization. The tracking phenomenon can be seen more clearly.



### Summary Measure Analysis of Longitudinal Data


```{r}

# Number of time steps
n <- RATS$Time %>% unique() %>% length()

# Summary data with mean and standard error of bprs by treatment and week 
RATSS <- RATS %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)


```



```{r fig_rats3, fig.path="figures/", fig.dim=c(7,5), results='hide', message=FALSE, warning=FALSE}

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  theme(legend.position = c(0.8,0.5)) +
  scale_y_continuous(name = "mean(weight) +/- se(weight)")

```

Mean weight profiles for the 3 groups in the RATS data. We can clearly see that group 1 is very different from other groups. The mean weight is significantly lower than in the other 2 groups. The difference is visible right from the start. So the actual weight profiles over time don't have much difference.

