# Chapter 3: Logistic regression

Data creation R script:   
https://github.com/tnissinen/IODS-project/blob/master/data/create_alc.R

### Loading and checking the structure of the data

```{r readdata_alc,echo=TRUE,results='hide',message=FALSE,warning=FALSE}
setwd("~/IODS-project")
library(dplyr)
library(tidyr)
library(ggplot2)
alc <- read.csv("~/IODS-project/data/alc.csv")

```

```{r datastructure_alc}
# structure of the data (all integers except gender that is "F"/"M")
colnames(alc)

glimpse(alc)
```

The dataset contains 382 observations with 35 variables

4 interesting variables:
activities - extra-curricular activities (binary: yes or no)
romantic - with a romantic relationship (binary: yes or no)
failures - number of past class failures (numeric: n if 1<=n<3, else 4)
studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)

My hypothesis
activities, romantic relationship and higher studytime will correlate negatively with alcohol consumption. Failures will probably correlate positively with alcohol consumption.


### Distributions and relatioships of the selected variables

Distributions 

```{r fig_alc2, fig.path="figures/", fig.dim=c(10,10), results='hide', message=FALSE, warning=FALSE}


alc_selected <- select(alc, one_of(c("activities", "romantic", "failures", "studytime")))
gather(alc_selected) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar()
```

Relationships with high alcohol usage


Activities
```{r}

g1 <- ggplot(data = alc, aes(x = alc_use, fill = activities))
g1 + geom_bar()

```

Romantic
```{r}

g1 <- ggplot(data = alc, aes(x = alc_use, fill = romantic))
g1 + geom_bar()

```

Failures
```{r}

qplot(alc_use, failures, data = alc) + geom_smooth(method = "lm")

```

Studytime
```{r}

qplot(alc_use, studytime, data = alc) + geom_smooth(method = "lm")

```


### Logistic regression for exploring the relationships

Summary of my explanatory variables coefficients:

```{r, results='asis'}

# find the model with glm()
m <- glm(high_use ~ failures + activities + studytime + romantic + failures, data = alc, family = "binomial")

# print out a summary of the model
results <- summary(m)

# print out a summary of the model
knitr::kable(results$coefficients, digits=3, caption="Coefficients")

```

We can see that my original hypothesis was correct. Although, only failures and studytime are statistically significant factors (p<0.05).


Odds ratios:

```{r, results='asis', message=FALSE}

# compute odds ratios (OR)
OR <- coef(m) %>% exp

# compute confidence intervals (CI)
CI <- confint(m) %>% exp

# print out the odds ratios with their confidence intervals
results <- cbind(OR, CI)

# print
knitr::kable(results, digits=3, caption="Odds ratios")

```


### Analysis of the predictive power of the model

```{r, results='asis'}

# predict() the probability of high_use
probabilities <- predict(m, type = "response")

# add the predicted probabilities to 'alc'
alc <- mutate(alc, probability = probabilities)

# use the probabilities to make a prediction of high_use
alc <- mutate(alc, prediction = probability > 0.5)

# tabulate the target variable versus the predictions
results <- table(high_use = alc$high_use, prediction = alc$prediction)

knitr::kable(results, digits=3, caption="Tabulate the target variable versus the predictions")
```

```{r fig_alc3, fig.path="figures/"}

g <- ggplot(alc, aes(x = probability, y = high_use, col = prediction))

# define the geom as points and draw the plot
g + geom_point()
# tabulate the target variable versus the predictions
tab <- table(high_use = alc$high_use, prediction = alc$prediction) %>% prop.table() %>% addmargins()


knitr::kable(tab, digits=3, caption="Tabulate the target variable versus the predictions (proportions)")
```

Total proportion of inaccurately classified individuals:
```{r}

v <- as.vector(table(high_use = alc$high_use, prediction = alc$prediction))
v2 <- (v[2]+v[3])/(v[1]+v[2]+v[3]+v[4])
v2

```

...
