# Chapter 2: Regression and model validation

Data creation R script:   
https://github.com/tnissinen/IODS-project/blob/master/data/create_learning2014.R

### 1. Loading and checking the structure of the data

```{r readdata,echo=TRUE,results='hide',message=FALSE,warning=FALSE}
setwd("~/IODS-project")
library(dplyr)

lrn14 <- read.csv("~/IODS-project/data/learning2014.csv")

```

```{r datastructure}
# structure of the data (all integers except gender that is "F"/"M")
str(lrn14)
```

The dataset contains 166 observations with 7 variables (gender ("M" = male/"F" = female), age, attitude toward statistics, avg. of deep learning grades, avg. of strategic learning grades, avg. of surface learning grades and total points).

### 2. Graphical overview of the dataset

```{r fig1, fig.path="figures/"}
pairs(lrn14)
```

```{r fig2, fig.path="figures/", fig.dim=c(10,10), results='hide', message=FALSE}
library(GGally)
library(ggplot2)

ggpairs(lrn14, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
```

    
From the graphical overview of the data I made following observations:

- the distribution of age variable shows that there are more young people (close to their 20s) in the dataset 
- the total points are quite evenly distributed over all ages, however the highest points are achieved by people below 40
- deep learning grades are higher than othe grades
- attitude seems to correlate heavily with points

```{r}
qplot(Attitude, Points, data = lrn14) + geom_smooth(method = "lm")
```

### 3. Linear regression

I chose Attitude, Age and Gender as explanatory variables:

```{r, results='asis'}
# fit a linear model
my_model <- lm(Points ~ Attitude + Age + gender, data = lrn14)

# print out a summary of the model
results <- summary(my_model)
knitr::kable(results$coefficients, digits=3, caption="Regression coefficients")
```

Age and Gender doesn't seem to have statistically significant relationship with points, so I removed them.

```{r, results='asis'}
# fit a linear model again with only Attitude
my_model <- lm(Points ~ Attitude, data = lrn14)

# summary of the model
results <- summary(my_model)

knitr::kable(results$coefficients, digits=3, caption="Regression coefficients")
```

### 4. Analysis of data and relationships

Attitude has an effect of about 0.35 to the total points with a standard error of about 0.057. Since the Pr value is close to zero, we can conclude that the attitude has a statistically significant relationship with the points.

Also Intercept is statistically significant in the model.

The R-squared statistic provides a measure of how well the model is fitting the actual data. In multiple regression settings, the R2 will always increase as more variables are included in the model. That’s why the adjusted R2 is the preferred measure as it adjusts for the number of variables considered

In this case the attitude has significant effect, but R squared value of 0.20 tells that it's clearly not the only factor. This can be also seen from the simple scatter plot of attitude vs. points.


### 5. Diagnostic plots

```{r fig3, fig.path="figures/"}
plot(my_model, which=c(1,2,5))
```

We assume that the error in our model is normally distributed. The Q-Q plot shows that the normality assumption seems to hold reasonably well.

The constant variance assumption can be evaluated from the Residuals vs fitted plot. There is not a very clear pattern visible, so we can conclude that the assumption holds quite well. However, the spread seems to decrease a little towards the far end of the residuals, which might be a small problem.

The leverage plot shows that some observations have a little more leverage than most, but the relative difference is still quite small (0.01 vs. around 0.04) 
